---
title: "Tipping points in sustainable agriculture? A coupled natural human systems approach to understanding diversified practices"
author:
  - name: Melissa Chapman
    affiliation: espm
  - name: Serge Wiltshire
    affiliation: espm
  - name: Timothy Bowles
    affiliation: espm
  - name: Claire Kremen
    affiliation: espm
  - name: Kathryn De Master
    affiliation: espm
  - name: "...others in group"
    affiliation: espm
  - name: Carl Boettiger
    affiliation: espm
    footnote: Corresponding Author
address:
  - code: espm
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: ubc
    address: "University of British Columbia"
abstract: |
keywords:
  - diversified practices,
  - ecosystem services
bibliography: alt_stable_states.bib
date: "`r Sys.Date()`"
output: rticles::elsevier_article
layout: 3p # review # review = doublespace, 3p = singlespace, 5p = two-column
header-includes: 
  - \journal{TBD}
  - \graphicspath{{figs/}}

---


```{r setup, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width=7, 
                      fig.height=3.5, 
                      fig.align = "center")
library(tidyverse)
library(MDPtoolbox)
library(mdplearning)
source("../R/continuous-model.R")
set.seed(123456)
```


```{r include = FALSE}
## We'll use these opinionated, typographic-centric-graphics
## This introduces a few additional system dependencies.  
## This whole block can be commented out for default graphics typography instead. 
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())


library(ggthemes)
library(patchwork) # devtools::install_github("/ggtext")
library(ggtext) # devtools::install_github("clauswilke/ggtext")
pal <- solarized_pal()(7)
scale_colour_discrete <- function(...) scale_colour_solarized()
scale_fill_discrete <- function(...) scale_fill_solarized()

## hrbrthemes needs cairo_pdf device
knitr::opts_chunk$set(dev = "cairo_pdf")

```


***Note: this is just me attempting a fresh outline to get a sense of what a greater CNHS-style pitch would look like.  Intending the outline to be just figures and topic-sentences of each paragraph / section for startes.***

# Introduction

- Agricultural systems are both important for their fundamental influence over human and ecosystem wellbeing, and a prime example of a global change issue in which societal and ecological issues are tightly enmeshed.

- Our understanding of these systems rarely reflects the dynamic and tightly integrated nature of societal and ecological processes in agricultural systems.  We advance a basic conceptual model

# Model setup



# Results

```{r model_def, results='hide'}
actions <- seq(0, 1, length = 100)
# Allow max(state) > max(actions) to avoid creating a hard boundary on ecosystem state
states <- seq(0, 1.5, length = 100)
params <- list(benefit = 1.57, cost = 1, sigma = 0.1, r = .1)
discount <- 0.97
transition_fn <- function(s, a, params) s + params$r * (a - s)
utility_fn <- function(s,a, params) params$benefit * s - params$cost * a
model <- continuous_model(states, actions, params, transition_fn, utility_fn)
```

```{r results = "hide"}
soln <- MDPtoolbox::mdp_value_iteration(model$P, model$U, discount)
soln_df <- tibble(state = states,
                  action = actions[soln$policy])
#write_csv(soln_df, "../data/soln_df.csv.gz")
```



```{r}
reps <- 500
init <- truncnorm::rtruncnorm(reps, 0, 1, 0.5, 0.2) %>% 
                  map_int(function(x) which.min(abs(x - states)))

sim <- function(soln, 
                Tmax = 10,
                x0 =  init){
  map_dfr(1:length(x0), 
          function(i)
                   mdp_planning(model$P, model$U, discount, 
                   policy = soln$policy,
                   x0 = x0[i], Tmax = Tmax), .id = "reps")
}

sims <- sim(soln)
#write_csv(sims, "../data/sims.csv.gz")
```

```{r }
sim_plot <- function(sims, panelA_opts = ggtitle(""), panelB_opts = ggtitle("")){
  
   df <- sims %>% 
    select(-obs,-value) %>% # tidy
    mutate(state = states[state], action = actions[action]) # rescale
  
   panelA <- df %>% 
    ggplot(aes(time, state, group = reps, col=time)) + 
    geom_path(alpha=0.1, show.legend = FALSE) + 
    panelA_opts
   
   Tmax <- max(sims$time)
   panelB <- df %>% filter(time %in% c(1, Tmax))  %>%
     ggplot() + geom_density(aes(state, group = time, fill = time), alpha=0.8) + 
     scale_fill_continuous(breaks=c(1,10)) +
     coord_flip() + 
     panelB_opts
   
  panelA + panelB + plot_layout(widths = c(3, 1))
}
```

```{r fig.cap="Simulations of the ecosystem state of 500 farming plots over time. Plots are initialized with a truncated normal distributed ecosystem state (mean state 0.5, standard deviation 0.2, truncated at 0,1).  Plots are then managed according to the decision rule as discussed in the main text."}
sim_plot(sims)
```




```{r fig_policyfn, fig.cap= "The emergent decision rule"}
soln_df %>%
  ggplot(aes(state,action)) + geom_point() +
  coord_cartesian(xlim = c(0,1))
```




```{r  results="hide"}
tenure_2 <- MDPtoolbox::mdp_finite_horizon(model$P, model$U, discount, N = 2)
sims_short_tenure <- sim(tenure_2)
#write_csv(sims_short_tenure, "../data/sims_short_tenure.csv.gz")
```

```{r fig_tenure, fig.cap="Under short land tenure (shown at 2 years), farmers on most farm plots opt to discontinue or not to adopt diversified practices. This results in a degredation of ecosystem state even among those plots with an initially high value. All parameters are as in previous figure, but decision problem is solved under the finite, 2-year horizon during which farmer will have access to the current plot and thus be able to reap the benefits of the ecosystem state."}
sim_plot(sims_short_tenure) 
```


## Subsidies

```{r results="hide"}
params_subsidy_burst <- list(benefit = 1.57, cost = 0, sigma = 0.1, r = .1)
params_subsidy_sustained <- list(benefit = 1.57, cost = .8, sigma = 0.1, r = .1)

## Note that farmers do not solve the finite-time horizon problem for the short period of the subsidy,
## because they are not "cashing out" (moving off the land) after the subsidy expires.
## This is still not the optimal solution, but is a reasonable approximation.
modelA <- continuous_model(states, actions, params_subsidy_burst, transition_fn, utility_fn)
solnA <- MDPtoolbox::mdp_value_iteration(modelA$P, modelA$U, discount)
  
## Do 2 years with large subsidy (no cost to adoption)
start <- sim(solnA, Tmax = 3, x0 = init)
x1 <- start %>% filter(time == 3) %>% pull(state)
## Go another 18 years with decision rule under no subsidy
rest <- sim(soln, Tmax = 18, x0 = x1) %>% filter(time!=1) %>% mutate(time = time+2)
simA <- bind_rows(start, rest)

## Do 10 years with minor subsidy (subsidize 10% of unit cost over first 10 years)
modelB <- continuous_model(states, actions, params_subsidy_sustained, transition_fn, utility_fn)
solnB <- MDPtoolbox::mdp_value_iteration(modelB$P, modelB$U, discount)
start <- sim(solnB, Tmax=10, x0 = init)  
x10 <- start %>% filter(time == 10) %>% pull(state)
## Go another 10 years with decision rule under no subsidy
rest <- sim(soln, Tmax = 10, x0 = x10) %>% filter(time!=1) %>% mutate(time = time+10)
simB <- bind_rows(start, rest)


#write_csv(simA, "../data/simA.csv.gz")
#write_csv(simB, "../data/simB.csv.gz")
```


```{r fig_subsidy, fig.height=7, fig.cap="Panel A: replicate simulations from 500 normally distributed starting states under a large subsidy (no direct cost to adoption) over two years. Panel B: the same starting conditions for 500 simulations under a smaller subsidy (cost = 0.9) for 10 years.  Ingoring discounting, subsidies have the same total cost After subsidy is removed, farmer adjusts their decision rule to that of no subsidy."}
sim_plot(simA, ggtitle("A. Large, short subsidy")) / 
   sim_plot(simB, ggtitle("B. Small, sustained subsidy"))
```


# Discussion


# Methods


# Acknowledgements

# References


# Appendix


{>> Where this bit winds up (and how concise it needs to be) will probably depend on the venue <<}

Mathematically, the farmer's decision model can be expressed as 

$$\max_{\lbrace a_t \rbrace} \mathbb{E} \left[ \sum_t^T U(x_t, a_t) \delta^t \right] $$

where $\lbrace a_t \rbrace$ is the set of available actions to be taken at each point in time $t$, $\delta$ is the discount rate, $\mathbb{E}$ the expectation operator, and $U(x_t, a_t)$ the utility which the farmer associates with being in state $x_t$ and taking action $a_t$ at time $t$. $T$ is the land tenure of the farm ($T = \infty$ if the farmer owns the land or otherwise expects to be able to farm the same land and thus benefit from the ecosystem services established there indefinitely)

We assume a simple model for the farmer's utility $U(x_t, a_t)$  as combination of the costs associated with adopting the diversified practice and the benefits derived from the ecosystem state (which is in turn influenced by the diversified practices or non-diversified practices adopted),

$$U(x_t, a_t) = b x_t  - c a_t$$
Where $x_t$ is the ecosystem state, $b$ the benefit associated, and $c a_t$ the cost of taking action $a_t$.  In general, more complicated nonlinear functions of both the ecosystem state and action are possible in this framework.  

The ecosystem state is also dynamic, evolving according to the transition function $f(x_t, a_t)$ 

$$x_{t+1} = f(x_t, a_t) := x_t + r \left(a_t - x_t \right)$$ 

This provides a minimal, one-parameter model in which the parameter $r$ sets the natural timescale at which the ecosystem can respond to a change in mangement practice.  

